---
title: "For loops"
author: ""
format: 
  html: default
code-annotations: select
---

## For Loops


Loops are a fundamental structure for repetition in programming.
`for` loops perform the same action for each item in a list of things.
The basic syntax is:

```{r}
#| eval: false
for (item in list_of_items) {
  do_something(item)
}
```

We can create a vector on the fly to loop a particular number of times:

```{r}
for (i in 1:5){
  print(i)
}
```

Or use a pre-existing vector or list. 

```{r}
volumes = c(1.6, 3, 8)
for (volume in volumes){
  mass <- 2.65 * volume ^ 0.9
  print(mass)
}
```

We also might want to loop over indices so we can access multiple vectors. 

```{r}
as <- c(2.65, 1.28, 3.29)
bs <- c(0.9, 1.1, 1.2)
volumes = c(1.6, 3, 8)
masses <- vector(mode="numeric", length=length(volumes))
for (i in 1:length(volumes)){
   mass <- as[i] * volumes[i] ^ bs[i]
   masses[i] <- mass
}
masses
```

We can use functions inside loops. For example, let’s take a function that returns an estimated mass if the `volume > 5` and `NA` if it’s not.

```{r}
est_mass <- function(volume, a, b){
  if (volume > 5) {
    mass <- a * volume ^ b
  } else {
    mass <- NA
  }
  return(mass)
}
```

We can then call the function to populate a vector item by item. 

```{r}
masses <- vector(mode="numeric", length=length(volumes))
for (i in 1:length(volumes)){
   mass <- est_mass(volumes[i], as[i], bs[i])
   masses[i] <- mass
}
masses
```

To note, this is the for loop equivalent of an `mapply` statement.

```{r}
masses_apply <- mapply(est_mass, volumes, as, bs)
masses_apply
```
In R we often want to use `apply` statements as opposed to explicitly writing loops. 
------------------------------------------------------------------------

*The materials in this lesson have been adapted from work created by the (HBC)\](http://bioinformatics.sph.harvard.edu/) and Data Carpentry (http://datacarpentry.org/), as well as materials created by Laurent Gatto, Charlotte Soneson, Jenny Drnevich, Robert Castelo, and Kevin Rue-Albert. These are open access materials distributed under the terms of the [Creative Commons Attribution license](https://creativecommons.org/licenses/by/4.0/) (CC BY 4.0), which permits unrestricted use, distribution, and reproduction in any medium, provided the original author and source are credited.*