---
title: "Exercise-2: Boxplot"
editor: visual
---

**Exercise 2: Boxplots**

A boxplot provides a graphical view of the distribution of data based on a five number summary:

-   The top and bottom of the box represent the (**1**) first and (**2**) third quartiles (25th and 75th percentiles, respectively).

<!-- -->

-   The line inside the box represents the (**3**) median (50th percentile).

-   The whiskers extending above and below the box represent the (**4**) maximum, and (**5**) minimum of a data set. The whiskers of the plot reach the minimum and maximum values that are not outliers.

> In this case, **outliers** are determined using the interquartile range (IQR), which is defined as: Q3 - Q1. Any values that exceeds 1.5 x IQR below Q1 or above Q3 are considered outliers and are represented as points above or below the whiskers.

1.  Boxplot

Generate a boxplot using the data in the `new_metadata` dataframe. Create a `ggplot2` code chunk with the following instructions:

1\. Use the `geom_boxplot()` layer to plot the differences in sample means between the Wt and KO genotypes.

2\. Use the `fill` **aesthetic** to look at differences in sample means between the celltypes within each genotype.

3\. Add a title to your plot.

4\. Add labels, 'Genotype' for the x-axis and 'Mean expression' for the y-axis.

5\. Make the following `theme()` changes:

-   Use the `theme_bw()` function to make the background white.

-   Change the size of your axes labels to 1.25x larger than the default.

-   Change the size of your plot title to 1.5x larger than default.

-   Center the plot title.

::: {.callout-tip appearance="simple"}
### Solution

```{r}

library(ggplot2)
load("data/new_metadata.RData")

ggplot(new_metadata) +
  geom_boxplot(aes(x = genotype, y = samplemeans, fill = celltype)) +
  ggtitle("Genotype differences in average gene expression") +
  xlab("Genotype") +
  ylab("Mean expression") +
  theme_bw() +
  theme(axis.title = element_text(size = rel(1.25))) +
  theme(plot.title=element_text(hjust = 0.5, size = rel(1.5)))

```
:::

2\. Changing the order of genotype on the Boxplot

Let's say you wanted to have the "`Wt`" boxplots displayed first on the left side, and "`KO`" on the right. How might you go about doing this?

To do this, your first question should be - How does `ggplot2` determine what to place where on the X-axis? The order of the genotype on the X axis is in alphabetical order. To change it, you need to make sure that the genotype column is a factor. And, the factor levels for that column are in the order you want on the X-axis.

1\. Factor the `new_metadata$genotype` column without creating any extra variables/objects and change the levels to `c("Wt", "KO")`

2\. Re-run the boxplot code chunk you created for the "Boxplot" exercise above.

3\. Changing default colors

::: {.callout-tip appearance="simple"}
```{r}
new_metadata$genotype <- factor(new_metadata$genotype, levels=c("Wt","KO"))
ggplot(new_metadata) +
  geom_boxplot(aes(x = genotype, y = samplemeans, fill = celltype)) +
  ggtitle("Genotype differences in average gene expression") +
  xlab("Genotype") +
  ylab("Mean expression") +
  theme_bw() +
  theme(axis.title = element_text(size = rel(1.25))) +
  theme(plot.title=element_text(hjust = 0.5, size = rel(1.5)))
```
:::

You can color the boxplot differently by using some specific layers:

1\. Add a new layer `scale_color_manual(values=c("purple","orange"))`.

-   Do you observe a change?

2\. Replace `scale_color_manual(values=c("purple","orange"))` with `scale_fill_manual(values=c("purple","orange"))`.

-   Do you observe a change?

-   In the scatterplot we drew in class, add a new layer `scale_color_manual(values=c("purple","orange"))`, do you observe a difference?

-   What do you think is the difference between `scale_color_manual()` and `scale_fill_manual()`?

3\. Back in your boxplot code, change the colors in the `scale_fill_manual()` layer to be your 2 favorite colors.

-   Are there any colors that you tried that did not work?

::: {.callout-tip appearance="simple"}
### Solution

```{r}
ggplot(new_metadata) +
  geom_boxplot(aes(x = genotype, y = samplemeans, fill = celltype)) +
  ggtitle("Genotype differences in average gene expression") +
  xlab("Genotype") +
  ylab("Mean expression") +
  theme_bw() +
  theme(axis.title = element_text(size = rel(1.25))) +
  theme(plot.title=element_text(hjust = 0.5, size = rel(1.5))) +
  #We can also use hex color values to choose colors
  scale_fill_manual(values=c("#F67280", "#355C7D"))
```
:::

*The materials in this lesson have been adapted from work created by the (HBC)\\\](http://bioinformatics.sph.harvard.edu/) and Data Carpentry (http://datacarpentry.org/). These are open access materials distributed under the terms of the \[Creative Commons Attribution license\](https://creativecommons.org/licenses/by/4.0/) (CC BY 4.0), which permits unrestricted use, distribution, and reproduction in any medium, provided the original author and source are credited.*
